On the Synchronization Power of the Openflow Data Plane

### 2015-04-19 inband-sync.tex
old iteration of stefan. now the new one starts :)

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions. We discuss different use cases
for our interface (consistent policy installation, load-balancing, consensus),
and also describe how to
perform basic forms of policy compositions in-band.
\end{abstract}

@@

\section*{todo}

stefan add old notes, bundle size complexity (how to make it minimal: store rule in one table, select it with a pointer / selector),
practical implementation (not even noviswitch has bundles)

we add all rules of the policy also! but we could add rules before and only activate later => set selector later (maybe set timeout); transaction size constant => keep bundle small important: packet atomicity so maybe on hold during bundle

@@

Being able to provide more complex and atomic sequences of transactions over this configuration
space can be appealing to resolve synchronization problems.
This paper investigates the implementation of a transactional interface 
(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.

\paragraph{Our Contributions}
This paper shows how to implement a very powerful transactional interface
 for software-defined networks. 
 Concretely, our implementation is based on
 standard Openflow~\cite{of-spec} primitives, which by themselves are very limited.
For example, we find that while Openflow's \emph{bundle} or \emph{rule overlap check}
features can serve as synchronization primitives,
without further mechanisms, their use is very limited:
with a bundle alone it is not possible to react to
and change switch configuration state in any non-trivial way (a bundle
essentially represents a single switch interaction),
and with the rule overlap check it is not possible to react to
update conflicts defined over non-overlapping flow spaces.
Moreover, we show that many powerful synchronous operations
can also be implemented without bundles.

The transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}.
For instance, a transaction allows to modify switch configurations in an interactive
and computational fashion, as well as to define complex notions
of conflicts.
Our implementation is based on distributed identifiers and makes use of two
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.
The claim, unclaim, and check-claim operations can be implemented individually without the Openflow
bundle feature.

Our paper therefore also provides the missing link for the read-modify-write object
postulated in~\cite{cpc}.

@@

Finally, we hope that our work can contribute to the ongoing discussion of what can be implemented
in-band in Openflow~\cite{compute,reclaim}, by taking a first look
at the relevant problem of distributed
synchronization.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
In Section~\ref{sec:motivation}, we discuss the limitations of interpreting
the Openflow bundle as a transaction, and motivate our approach.
We then present our approach in detail in Section~\ref{sec:main}, and discuss use cases in Section~\ref{sec:apps}.
%Section~\ref{sec:compo} describes how to perform in-band composition.
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}. Some technical details are postponed
to the Appendix.

@@

The synchronization mechanisms presented in this paper
are based on Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.

In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules (flow entries). Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline; during the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information. Concretely, instructions can be
used to define additional tables to be visited (goto instruction),
to modify the set of to-be-applied actions (either by appending,
deleting, or modifying actions), or immediately apply some
actions to the packet. A meta-data field can be used to exchange
information between tables; it is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

\liron{I think that maybe the rest of this section should be explained by the following subsections, providing in depth description of the OpenFlow features.}
\liron{Another direction is to have a simplified explenation as now and to refference spesific sections of the standard in side our paper.}

\subsection{The FlowMod Command}

\emph{FlowMod} is the main messages used in Openflow 
to update the switch configuration. 
In a nutshell, \emph{FlowMod} specifies a new flow entry or modifications to existing flow entries.
A flow entry is mainly composed from a match defined by a ternary patterns (a value and a mask) over packet header fields, and from actions.

%,a match and an action. 
The standard behavior is that the new flow entry simply replaces an existing one
with exactly the same match, and is installed
in addition to an existing one if the match only
partially overlaps. The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority. \stefan{liron, is this correct?}\liron{a bit not clear, lets talk about this}

\emph{FlowMod} can be used together with a the CHECK\_OVERLAP flag
of \emph{FlowMod}, in order to detect conflicts: if the flow spaces of an old
and the new rule overlap, 
that is, if there is an entry with overlapping match part but different action, 
 and if the two rules have the same priority, then 
\emph{FlowMod} fails.

As we will see, in order to implement our synchronization objects,
the \emph{FlowMod} and CHECK\_OVERLAP features are sufficient.

\subsection{Bundling}

The Openflow bundle can be seen as a ``mini-transaction'':
a bundles aggregates multiple control commands (e.g., FlowMod messages)
into one execution, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
This bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.
If used together with the CHECK\_OVERLAP flag, the bundle command also returns a reason 
if the modification failed: the failed command in the bundle:
a CHECK\_OVERLAP flag can be defined for each command in the bundle.
\liron{these last two sentences are not cleaar}

Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the xid of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

We will make use of bundling only for policy updates. 
Bundling is \emph{not required} to implement transactions.

In the remainder of this paper, we will use the notation 
$\textbf{bundle}\{\;\ldots\;\}$ to refer to represent a
sequence of commands in a bundle.
\stefan{liron: i think the following needs more details $:-)$ :} A bundle will always have the following structure:
it first contains a \emph{send message} to start bundle,
followed by a sequence of messages wrapped in bundle and bundle ID, 
and finally end bundle control message for commit.
A bundle always returns a binary result: \emph{error} or \emph{success}.
\liron{the notion of transaction presented next is confusing considering the bundle just explained}
Concretely, a transaction can represent between two to five switch interactions:
if limited space, Algo 7: read, claim id, read again, transaction, unclaim;
if if unlimited space: alg 6 is unbounded, increase policy ID by one.
transactions = five switch interactions , claim pid (line 5+6) together in bundle => one iteraction less?

In general, it is desirable to minimize the number of transactions \stefan{stefan should cite \cite{Jin2014Dionysus} here}

To commit the transaction,  the controller sends a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_COMMIT\_REQUEST} and waits until a matching control
message from the switch is received. When the switch receives the
commit request it tries to atomically process all the buffered messages
equipped with the corresponding bundle identifier and, if the
processing is successful for all of them, returns  a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_CLOSE\_REPLY}; otherwise it sends a specific error
message  pointing out the aborted transactional operation.

Algorithm~\ref{alg:commit} describes how to commit
the bundle. 
\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \textsf{OFPBCT\_COMMIT\_REQUEST}
    		\State  $cmd.flags \gets $ \textsf{OFPBF\_ATOMIC}
    		\State send $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.35\columnwidth]{loadbal-before.pdf}~\includegraphics[width=0.35\columnwidth]{loadbal.pdf}~~
\includegraphics[width=0.8\columnwidth]{pic.pdf}\\
\caption{\emph{Left:} Load-balancing gone wrong. \emph{Right:} Illustration of transactional abstraction which  allows
controllers to modify switch configurations in complex ways,
and
depending on other flow spaces. Colors indicate different actions. \stefan{lets make two real transactions: one bundle and one 2 or 5 interactions?}}\label{fig:illu}
\end{figure*}

Finally, we remark that our solution is only based on flow tables
and does not make use
of group tables. 
Moreover, we do not make use of cookies: rules
are always identified by exact matches.

@@

\stefan{liron, maybe we can discuss this section together: we need some more ideas}

Before presenting our transactional interface and implementation,
let us first discuss the limitations of the Openflow features discussed above.
At first sight, the bundle feature seems to provide everything
we need: it defines a sequence of read and write operations
 on the switch state which are executed atomically.
 However, such bundle transactions come with several
 limitations, as discussed in the following. Moreover,
 as we will see, our mechanism also does not \emph{need}
 the bundle feature.

A bundle does not allow
 to change the switch state in any non-trivial way,
 as the read and write actions in the bundle do
 not support any conditional modifications or computations which
 depend on the read results. By leveraging
 the check overlap feature, it is at least possible
 to abort transactions depending on certain conditions
 in the overlapping
 flow spaces. However, in many cases, for example
 a load-balancing application,
 it can be interesting to define more general
 dependencies, which depend on, from a logical perspective,
 seemingly
 completely independent flow spaces,
 which however share the underlying physical network.
In other words, the bundle feature only allows us to
define ``one-shot transactions'',
transactions which do not support any further controller interactions.
Given
the limited computational options the Openflow interface of a switch offers,
it is difficult to implement more complex behavior.

\textbf{Example 1.} To give an example, let us consider a controller application in charge
of \emph{load-balancing}. The objective of the load-balancer
is to distribute flows equally across links or servers.
If multiple controllers concurrently and independently
update forwarding rules, they may unvoluntarily
unbalance the flow allocation. Note that this
is true independently of whether the flow spaces
affected by the respective controllers overlap or not.
More concretely, assume that two controllers concurrently
decide to forward traffic $1*$ and $0*$ to the same server.
Flow space overlap checks cannot be used to detect
such conflicts.
FIXME: make this example better

\begin{figure}[ht]
\centering
\includegraphics[width=1\columnwidth]{no-sync.png}\\
\caption{}\label{fig:no-sync}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\columnwidth]{sync-with-cas.png}\\
\caption{}\label{fig:cas-sync}
\end{figure}

\textbf{Example 2.} To give another example,
consider the problem of modifying previously
installed forwarding policies.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check, to abort the transaction if potentially conflicting
rules were already there), it cannot be used
to modify rules in non-trivial ways.
In other words, the bundle and overlap check only helps
for the conditional creation of rules, not their modification.

\stefan{It would be cool if we could make these examples more concrete,
at least one, and get back to it later again!}

%\section{The Container Object}\label{sec:vision}
\section{Transactional Synchronization}\label{sec:main}

This section describes how to introduce more powerful transactions.
Our implementation is based on two new synchronization objects,
allowing to coordinate controllers using identifiers.
At a high level, we export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.

@@

We implement transactional access to  two kinds of objects: a read-write \emph{memory} abstraction
object, which we later use for storing the current used policy identifier, and a
\emph{resource-claimer} object, provides claiming a resource id (e.g. policy \emph{identifier}) in a
given (bounded or unbounded) set.
In theory one can use the memory abstraction to implement the resource claimer, 
however we found it far more simple and efficient to consider the resource claimer as base object.
%

A memory can be accessed with the following sequential
semantics:

\begin{itemize}
\item By performing $\textit{Write}(addr,k)$,
  $addr,k\in\Nat$, the value at memory offset $addr$ is set to $k$. For
  example, the offset $addr$ identifies the policy (identifier), 
  but our approach is more general.

\item $\textit{\compare}(addr,k)$, $addr,k\in\Nat$,
aborts current transaction if the value at memory offset $addr$ is not equal to $k$.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).

Note that reading the memory can be done by just reading the switch configuration however such action doesn't affect the execution of the transaction (as $\textit{\compare}$ does).
As we will see, this abstraction can be implemented
without using the notion of bundles.

An immediate consequence of this memory interface is 
that we can implement a \emph{compare-and-set} (CAS) operation that takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\textit{true}$; otherwise, $\textit{false}$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort if
the transaction invoking it).

A resource-claimer object implements a weak type of locking, exports operations \emph{claim},
\emph{unclaim} and \emph{\claimcheck} with the following sequential
semantics:
%The container synchronization object introduced in this paper
%is essentially a transactional memory abstraction which can be implemented
%in standard Openflow switches.
%Before presenting the technical details in the next section,%
%we give a brief overview.

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the Openflow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item by performing $\textit{claim}(k)$, $i\in[N]$,
  $k\in\Nat$, controller $i$ \emph{claims}
  identifier $k$;

\item by performing $\textit{unclaim}(k)$, $i\in[N]$, $k\in\Nat$, controller $i$
  \emph{unclaim} identifier $k$;

\item $\textit{\claimcheck}(k)$, $k\in\Nat$,
abort current transaction if $k$ is currently claimed by any controller.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).

@@

\subsubsection{A memory abstraction}

>>>>

\newcommand{\FlowMod}{\emph{FlowMod}\xspace}
\newcommand{\match}{\emph{match}\xspace}
\newcommand{\action}{\emph{action}\xspace}
\newcommand{\flags}{\emph{flags}\xspace}
\newcommand{\checko}{\texttt{chkov}\xspace}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions. 
We show how our interface can be used for applications such as 
in-band controller consensus, consistent policy installation, or load-balancing.
\end{abstract}

@@

%
%
%\section*{todo}
%
%stefan add old notes, bundle size complexity (how to make it minimal: store rule in one table, select it with a pointer / selector),
%practical implementation (not even noviswitch has bundles)
%
%we add all rules of the policy also! but we could add rules before and only activate later => set selector later (maybe set timeout); transaction size constant => keep %bundle small important: packet atomicity so maybe on hold during bundle
%
%

@@

By providing more complex and atomic transactions over the
switch configuration
space, several synchronization problems can be resolved.
This paper investigates the implementation of a transactional interface
(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.

\paragraph{Our Contributions}
This paper shows how to implement a powerful transactional interface
for software-defined networks.
This transactional interface is reminiscent of
classic transactional memory abstractions~\cite{stm-st95,tm-book}
and
can be used to resolve
synchronization problems of distributed control planes efficiently. 
We show that this transactional interface can be used to implement
consensus among distributed controllers or to consistently install a policy (also providing 
the missing link for the read-modify-write object
postulated in~\cite{cpc}).
The atomic transactions supported by our interface
may also define complex notions
of conditions when transactions stand in conflict and should be serialized:
For instance, conflicts may be defined for transactions which seemingly
operate over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is attractive,
e.g., for a load-balancing application.

Interestingly, our transactional interface can be implemented using standard Openflow~\cite{of-spec} primitives, 
which by themselves are of limited use.
Concretely, we introduce, and implement in standard Openflow,
two new synchronization objects for managing \emph{distributed identifiers}:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.

We hope that our work can also contribute to the ongoing discussion of what can be implemented
in-band in Openflow~\cite{compute,reclaim}, by taking a first look
at the relevant problem of distributed
synchronization.

@@

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
In Section~\ref{sec:motivation}, we discuss the limitations of the Openflow
primitives and motivate the need for transactions.
We then present our approach in detail in Section~\ref{sec:main}, and discuss use cases in Section~\ref{sec:apps}.
%Section~\ref{sec:compo} describes how to perform in-band composition.
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}. Some technical details are postponed
to the Appendix.

@@

The transactional interface presented in this paper can be implemented in 
Openflow~\cite{of-spec}, the standard SDN protocol today.
% In this section,
%we provide the necessary background.
In a nutshell,
Openflow
follows a match-action concept: Openflow switches store
rules (installed by the controller) consisting of a match and an
action part. A packet matched by a certain rule will be subject
to the associated action (to be immediately applied now or later).
For example, an action can define a port to which the
matched packet should be forwarded, or add or change a tag
(a certain part in the packet header).
Each Openflow switch stores one or multiple flow tables,
each of which contains a set of rules: the \emph{flow entries}.

Flow
tables form a pipeline, and flow entries are ordered according
to priorities: A packet arriving at a switch is first checked by
the rule of the highest priority in table 0: the fields of the
data packet are compared with the match fields of that rule,
and if they fit, some instructions (the actions) are executed;
subsequently, lower priority rules are checked. Depending on
the outcome of the table 0 processing, the packet may be sent
to additional flow tables in the pipeline. 
Instructions can be
used to define additional tables to be visited (a ``goto instruction''),
to modify the set of actions to be applied later (either by appending,
deleting, or modifying actions), or to immediately apply some
actions to the packet.

During the traversal of
the flow tables, the packet can be equipped with meta-tags to
transfer temporary information between tables.
The meta-data field is part of the header that can be
inserted or removed from a packet via push and pop actions.
In general, a packet can be matched against any of its header
fields, and fields can be wildcarded and sometimes bitmasked.
(For instance, the meta-data field is maskable.) If no rule
matches, the packet is dropped.

%\liron{I think that maybe the rest of this section should be explained by the following subsections, providing in depth description of the OpenFlow features.}
%\liron{Another direction is to have a simplified explenation as now and to refference spesific sections of the standard in side our paper.}

\paragraph{The FlowMod Command}
The main messages used in Openflow
to update the switch configuration is called \emph{FlowMod}.
In a nutshell, \emph{FlowMod} either specifies a new flow entry or 
modifications to existing flow entries.
A flow entry consists of: (1) a match defined by a ternary pattern over packet header fields
(a value and a mask), and (2) a set of actions.
%,a match and an action.
The standard behavior is that the new flow entry will simply \emph{replace} 
an existing one if the existing one has exactly the \emph{same match}, 
but will be installed
\emph{in addition} to existing entries if they only
partially overlap (or define flowspace subsets or supersets). 
The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority.

In order to avoid conflicts, \emph{FlowMod} can be used together with a the $\checko$ flag ($CHECK\_OVERLAP$):
if the flow spaces of an old
and the new rule overlap,
that is, if there is an entry with an overlapping match part but a
different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.

We will exploit the \emph{FlowMod} command in combination with
$\checko$ to implement synchronization objects which form
the basis of our transactional interface.
We will use the following notation to create a \emph{FlowMod} message:
$\FlowMod(\match, \action, \flags)$, $\match$ denotes the match,
$\action$ the action, and $\flags$ the flags to activate the $\checko$ feature.
The \emph{FlowMod} message (and also the flow entries) include more parameters,
which however are irrelevant for our implementation: we suggest to
use the default values.

\paragraph{Bundling}
Another interesting feature in Openflow 
is the \emph{bundle}. A bundle 
aggregates multiple control commands (e.g., \emph{FlowMod} messages)
into one message, with \emph{all-or-nothing semantics}: 
either all of the commands are
performed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled commands cannot
be performed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
When properly configured,
the bundled processing can be done in a
\emph{packet-atomic} way, in the sense that 
each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
bundle (i.e., all its commands) took effect.
If multiple \emph{FlowMod} messages with the $\checko$ flag are sent as a single bundle, 
the bundle command also returns a reason
why the overlap checks failed: it returns the failed command in the bundle.
Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the \stefan{what is this:} \emph{xid} of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

%:a $\checko$ flag can be defined for each command in the bundle.\liron{these last two sentences are not cleaar}

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns 
$\top$ if and only if 
all commands are performed successfully, and otherwise $\bot$.
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \textsf{OFPT\_BUNDLE\_CONTROL} message of type
\textsf{OFPBCT\_OPEN\_REQUEST}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an 
\emph{add to bundle message} (\textsf{OFPT\_BUNDLE\_ADD\_MESSAGE}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\textsf{OFPT\_BUNDLE\_CONTROL}) of 
type \textsf{OFPBCT\_COMMIT\_REQUEST}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

%\liron{the notion of transaction presented next is confusing considering the bundle just explained}
%Concretely, a transaction can represent between two to five switch interactions:
%if limited space, Algo 7: read, claim id, read again, transaction, unclaim;
%if if unlimited space: alg 6 is unbounded, increase policy ID by one.
%transactions = five switch interactions , claim pid (line 5+6) together in bundle => one iteraction less?

%In general, it is desirable to minimize the number of transactions \stefan{stefan should cite \cite{Jin2014Dionysus} here}

\hide{
To commit the transaction,  the controller sends a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_COMMIT\_REQUEST} and waits until a matching control
message from the switch is received. When the switch receives the
commit request it tries to atomically process all the buffered messages
equipped with the corresponding bundle identifier and, if the
processing is successful for all of them, returns  a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_CLOSE\_REPLY}; otherwise it sends a specific error
message  pointing out the aborted transactional operation.

Algorithm~\ref{alg:commit} describes how to commit
the bundle.
\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \textsf{OFPBCT\_COMMIT\_REQUEST}
    		\State  $cmd.flags \gets $ \textsf{OFPBF\_ATOMIC}
    		\State send $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}
}

@@

%\begin{figure*}[t]
%\centering
%\includegraphics[width=0.35\columnwidth]{loadbal-before.pdf}~\includegraphics[width=0.35\columnwidth]{loadbal.pdf}~~
%\includegraphics[width=0.8\columnwidth]{pic.pdf}\\
%\caption{\emph{Left:} Load-balancing gone wrong. \emph{Right:} Illustration of transactional abstraction which  allows
%controllers to modify switch configurations in complex ways,
%and
%depending on other flow spaces. Colors indicate different actions. \stefan{lets make two real transactions: one bundle and one 2 or 5 interactions?}}\label{fig:illu}
%\end{figure*}

Finally, we remark that our solution is only based on flow tables
and does not make use
of \emph{Openflow group tables}, nor do we make use of \emph{cookies}: rules
are always identified by exact matches.

@@

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface: the bundle
 defines a sequence of read and write operations
 on the switch state which are executed atomically.
 However, while the bundle can indeed be seen 
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the read and write actions in the bundle do
 not support any conditional modifications or computations which
 depend on the read results. 
But also using the bundle together with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} transactions depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which transactions should not take effect. Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

\begin{enumerate}
\item \textbf{Example 1: More complex dependencies.} Let us consider a controller application in charge
of \emph{load-balancing}. 
In a load-balancing application, seemingly independent actions, defined over
completely independent logical flow spaces,
may in fact be dependent due to collocation in the underlying physical network.
To be more concrete, say that the objective of the load-balancer
is to distribute flows equally across links or servers.
If multiple controllers concurrently and independently
update forwarding rules (say, two ``independent'' flows with source IP address ranges $1*$ and $0*$), 
they may unvoluntarily
unbalance the flow allocation, as the managed flows are embedded
on the same physical links.  
Flow space overlap checks cannot be used to detect
such conflicts.

\stefan{todo: explain figure}

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding policies.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be used
to modify rules in non-trivial ways.
In other words, the bundle and overlap check only helps
for the conditional creation of rules, not their modification.
\end{enumerate}

\begin{figure*}[t]
\centering
\includegraphics[width=1\columnwidth]{no-sync.png}~\includegraphics[width=1\columnwidth]{sync-with-cas.png}\\
\caption{\emph{Left:} No synchronization. \emph{Right:} CAS.}
\end{figure*}

As we will see, we do not even need the bundle feature
to implement the powerful synchronization objects of our transactional interface.
However, the bundle feature can still be useful in the context of
consistent policy updates.

%At first sight, a bundle seems to be a powerful concept
%for the implementation of a transactional interface. However,
%while the  bundle can indeed be seen as a ``mini-transaction'',
%in itself it is not very useful. In fact, as we will discuss
%in more detail in the next section, we will not make
%use of the bundle in the implementation of our synchronization
%objects, but will employ it for our case study
%of consistent policy updates.%
%
%Before presenting our transactional interface and implementation,
%let us first discuss the limitations of the Openflow features discussed above.

%\section{The Container Object}\label{sec:vision}
\section{Transaction $\&$ Synchronization}\label{sec:main}

Given this motivation, we now present a more powerful transactional interface.
At a high level, we aim to export a
transactional interface which allows each controller $i\in[N]$ ($N\geq
2$) to perform multiple
operations on the configuration of an OpenFlow switch as an atomic
\emph{transaction}, inspired by the transactional-memory paradigm~\cite{stm-st95,tm-book}. A transaction may \emph{commit}, in
which case it appears as executed sequentially, or \emph{abort}, in
which none of its operations takes effect.

Our approach is based on the distributed management of \emph{identifiers (IDs)}. In particular, 
we introduce two new synchronization objects (and describe their implementation in standard Openflow),
which allows controllers to coordinate 
themselves based on these identifiers.

@@

We implement transactional access to  two kinds of objects: 
a read-write \emph{memory} abstraction
object, which we will use for storing the current used policy identifier, and a
\emph{resource-claimer} object, which allows controllers to claim a resource ID, 
e.g., a policy \emph{identifier} in a
given (bounded or unbounded) set.
%While In theory one can use the memory abstraction to implement the resource claimer,
%however we found it far more simple and efficient to consider the resource claimer as base object.
%

With these base objects, we provide the following shared memory abstraction, with 
write-compare access semantics:
\begin{itemize}
\item By performing $\textit{Write}(addr,k)$,
  $addr,k\in\Nat$, the value at memory offset $addr$ is set to $k$. For
  example, the offset $addr$ may be used to identify the policy (identifier),
  but our approach is more general.

\item The operation $\textit{\compare}(addr,k)$, $addr,k\in\Nat$,
aborts current transaction, if the value at memory offset $addr$ is not equal to $k$.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).
\end{itemize}

As we will see, the bundle feature is not required to implement this abstraction.
Also note that memory read operations correspond to reading the switch 
configuration, and are simple as they do not affect the execution of the transaction 
(unlike the $\textit{\compare}$ operation).

With this memory interface we can immediately implement 
a powerful \emph{compare-and-set} (CAS) operation which takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\top$; otherwise, $\bot$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort if
the transaction invoking it).

Our resource-claimer object implements a weak type of locking, and exports operations \emph{claim},
\emph{unclaim} and \emph{\claimcheck} with the following sequential
semantics:
%The container synchronization object introduced in this paper
%is essentially a transactional memory abstraction which can be implemented
%in standard Openflow switches.
%Before presenting the technical details in the next section,%
%we give a brief overview.

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the Openflow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item With $\textit{claim}(k)$, $i\in[N]$,
  $k\in\Nat$, a controller $i$ \emph{claims}
  identifier $k$.

\item With $\textit{unclaim}(k)$, $i\in[N]$, $k\in\Nat$, controller $i$
  \emph{unclaims} identifier $k$;

\item If $k$ is currently claimed by any controller, 
$\textit{\claimcheck}(k)$, $k\in\Nat$,
aborts the current transaction.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).

@@

\paragraph{A memory abstraction}
